---
title: "semdistflow tutorial"
author: "Celia Litovsky, PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    css: !expr ("/Users/tuo17460/Google Drive/StylesTemplates/jr_15.css")
---

```{r setup, include=FALSE}
library(semdistflow)
library(tidytext)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(stringr)
library(zoo)
library(stringi)
library(printr)
library(gutenbergr)
library(readr)
library(here)
```

# Preparing Your Texts

Each text you wish to process should be saved as a .txt file. Your texts should be stored in one folder. The file names in a given folder should be unique. 

# Loading your texts
```{r}
mytexts <- readtxt(here("mytexts")) # input path name to .txt files, read in files
```

# Read in text
```{r}
### You can input text in one of three main ways: (1) from Project Gutenberg, (2), from .txt file(s), or  (3) from directly typing it in

## text from Project Gutenberg (requires gutenbergr package)
janeeyre <- gutenberg_download(1260, mirror = "http://mirrors.xmission.com/gutenberg/")

# # need to change column names for cleaning script, new_name = old_name
# janeeyre %>%
#   rename(doc_id = gutenberg_id,
#          doc_text = text) -> janeeyre

# change doc_id to something meaningful for later
# janeeyre %>%
#   mutate(doc_id = "janeeyre") -> janeeyre

# remove gutenberg_id column
janeeyre <- subset(janeeyre, select=text)

# remove empty rows
janeeyre <- janeeyre[!apply(janeeyre == "", 1, all),]

# remove extra spaces
janeeyre <- trimws(janeeyre)

write.table(janeeyre, "janeeyre_full.txt", row.names = FALSE, col.names = FALSE)

# read the Gutenberg txt file back in
janeeyre <- readtxt("/Users/litovsky/Dropbox/mytexts/")

## from .txt file - give the function readtxt the path name to a folder
cats <- readtxt("/Users/litovsky/Dropbox/COG Workshop 2022 semdistflow/")
cats %>%
  mutate(doc_id = "cats") -> cats

## or just type in your text here
doc_id <- "fox" # each document will have its own doc_id
doc_text <- "The quick brown fox jumps over the lazy dog."
fox <- as.data.frame(cbind(doc_id,doc_text))
```

The resulting data.frame should have two columns "doc_id" and "doc_text". 
```{r}
print(janeeyre)
print(cats)
print(fox)
```

# "Cleaning" your data 
```{r clean,}
janeeyre.clean <- cleanme(janeeyre, replace_pronoun = TRUE) # performs all cleaning functions on the texts in mytexts. replace_pronoun = TRUE changes pronouns to simple antecedent (e.g., he -> person). replace_pronoun = FALSE keeps pronouns as-is
cats.clean <- cleanme(cats, replace_pronoun = TRUE)
fox.clean <- cleanme(fox, replace_pronoun = TRUE)
```

# Notice new column "doc_clean" has been added with the cleaned text

```{r, clean table print, echo=FALSE}
head(janeeyre.clean)
head(cats.clean)
head(fox.clean)
```

# Computing cosine distances
```{r clean,}
janeeyre.dist <- distme(janeeyre.clean)
cats.dist <- distme(cats.clean)
fox.dist <- distme(fox.clean)
```

